\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2020-12-11 icepack.tex
Add funding info
More grants acknowledgment
Add corrections from Ian

\Author[1]{Daniel}{Shapero}
\Author[2]{Jessica}{Badgeley}
\Author[2]{Andrew}{Hoffmann}

@@

\runningauthor{Shapero, Badgeley, Hoffmann}

@@

\begin{abstract}
We introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
Icepack is built on the finite element modeling library Firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
The diagnostic models in icepack are formulated through action principles that are specified in UFL.
The components of each action functional can be substituted for different forms of the user's choosing, which makes it easy to experiment with the model physics.
The action functional itself can be used to define a solver convergence criterion that is independent of the mesh and requires little tuning on the part of the user.
Icepack includes the 2D shallow ice and shallow stream models.
We have also defined a 3D hybrid model based on spectral semi-discretization of the Blatter-Pattyn equations.
Finally, icepack implements a Gauss-Newton solver for inverse problems that runs substantially faster than the standard BFGS method used in the glaciological literature.
The overall design philosophy of icepack is to be as usable as possible for a wide a swathe of the glaciological community, including both experts and novices in computational science.
\end{abstract}

@@

Numerical modeling has become an essential part of the workflow of glaciologists across all disciplines.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item predicting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood,
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like.
\end{enumerate}
Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several glacier flow models already exist and are effective in the hands of experts.
These models are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
Many researchers in glaciology, however, receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
Our goal is to make a tool that will be both accessible to non-experts and useful for experts.

The glacier flow modeling package closest in spirit to icepack is VarGlaS \citep{brinkerhoff2013data}.
VarGlaS is implemented using the finite element modeling package FEniCS \citep{logg2012automated}, whereas icepack is built on top of Firedrake.
The Firedrake project began as an outgrowth of FEniCS.
Both packages share a similar goal of saving users from manually writing low-level code for assembling the systems of equations that discretize their physics model.
Instead, users describe the weak form of the partial differential equations they wish to solve using a high-level domain-specific language (DSL) called the \emph{Unified Form Language} or UFL \citep{alnaes2014unified}.
This DSL is embedded entirely into Python, i.e. the complete syntax of UFL can be mapped directly onto overloaded operators in the Python programming language.
Both FEniCS and Firedrake then generate optimized C or C++ code to assemble the discretized system of equations from this symbolic description of the problem \citep{kirby2006compiler, rathgeber2016firedrake}.
Combining a DSL and a code generator frees users from the very error-prone process of writing these assembly kernels themselves, and makes the syntax of the code align closer to the syntax of written mathematical expressions.
Additionally, having a high-level symbolic description of the problem makes it possible to automatically derive tangent linear models and adjoints \citep{mitusch2019dolfin}.

Icepack improves upon the groundwork laid in VarGlaS in three main respects.
First, icepack includes a simple 3D flow model that uses several features only available in Firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated}.
This model will be described in section \ref{sec:physics-hybrid-model}.
Second, the architecture of icepack is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of the models included.
The software design that made this possible is discussed in section \S\ref{sec:physics-substitution}.
Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS; see section \S\ref{sec:numerics-inverse-solvers}.

The two main components of a glacier flow model are a \emph{diagnostic} and a \emph{prognostic} equation.
The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation.
The inputs to the diagnostic equation are the ice thickness, surface elevation, velocity at the inflow boundary, rheology, and bed friction coefficient.
The output of the diagnostic equation is the ice velocity throughout the entire domain.
The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
The inputs to the prognostic equation are the current value of the ice thickness, velocity, and the surface and basal mass balance.
The output is the ice thickness at the subsequent timestep.
Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation.

The rheology and friction coefficient are functions of other fields that have their own evolution equations.
For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock.
The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.

@@

@@

The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
The prognostic model is a conservative advection equation:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
The thickness is represented using continuous, piecewise polynomial basis functions in each cell of the mesh.
The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
The current prognostic model implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
We have not yet implemented a solver that works for discontinuous basis functions, but this extension is completely feasible within our framework and may come in a future release of icepack.

The surface elevation is calculated at each timestep of a simulation as
\begin{equation}
    s = \max\{b + h, (1 - \rho_I / \rho_W)h\},
\end{equation}
the first case corresponding to grounded ice and the second case corresponding to floating ice.
In the immediate vicinity of the grounding line, the assumption that the floating ice is in hydrostatic equilibrium with the ocean fails.
Most models assume hydrostasy and icepack does as well.
Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.
We have not implemented this feature but support for non-hydrostatic ice shelves may come in a future release of icepack.

@@

There are four \emph{diagnostic} or \emph{momentum transport} models implemented in icepack.
For each of the diagnostic models, we use a formulation of the physics based on a \emph{minimization} or \emph{action principle} \citep{dukowicz2010consistent}.
Action principles are completely equivalent to the usual weak form of a partial differential equation but have certain numerical advantages that we will describe below.

The most complex and most physically accurate model is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
(In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
The only approximation that the first-order model makes is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.

From the first-order model, two approximations are possible.
First, the \emph{shallow ice approximation} (SIA) comes from assuming that vertical shear is the dominant mode of ice flow.
The SIA model \citep{hutter1981effect} describes the interior of ice sheets well, where flow can be simply described as bed-parallel shear and surface and basal slopes are small.
This approximation breaks down near ice divides, where vertical velocities dominate, and near ice margins, where basal sliding can be a large fraction of the surface speed.
Second, one could assume that ice flow is purely by horizontal extension and that the surface and bed velocities are practically the same.
This is called the \emph{shallow stream approximation} (SSA).
The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.
Finally, for floating ice, the shallow stream approximation simplifies further to the \emph{shallow shelf approximation}.
The shallow shelf model has no bed friction and the ice surface elevation can be eliminated from the system.

All the diagnostic models inherit a fundamental nonlinearity in the mechanics of ice flow.
For a Newtonian viscous fluid, the stress tensor $\tau$ and the strain rate tensor $\dot\varepsilon$ are linearly proportional to each other.
Glaciers, however, have a nonlinear constitutive relation:
\begin{equation}
    \dot\varepsilon = A|\tau|^{n - 1}\tau
\end{equation}
where $A$ is the \emph{fluidity} and $n$ is the \emph{Glen flow law exponent} \citep{cuffey2010physics}.
The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.

@@

The action principle is especially useful for designing a robust numerical solver.
For viscous flow problems near to steady-state, the action is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge from any initial guess.
Minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation, although the two formulations are equivalent.

@@

The shallow ice model can be derived from the Blatter-Pattyn approximation by assuming the ice flow is dominated by bed-parallel shear and that surface and basal slopes are small, i.e. $\partial u/\partial x \ll \partial u/\partial z$.
The result is a 2D system of equations for depth-averaged horizontal velocity.
The class in icepack that represents this physics model is called \texttt{ShallowIce}.
The terms in the action functional are:
\begin{align}
    \text{mass} & =\frac{1}{2}\int_\Omega u\cdot u\,dx\\\
    \text{gravity} & = \int_\Omega\frac{2A(\rho_I g)^n}{n+2} h^{n + 1}|\nabla s|^{n - 1}\nabla s\cdot u\,dx\\
    \text{penalty} & = \frac{1}{2}\int_\Omega \ell^2\nabla u\cdot \nabla u\,dx
\end{align}
and the action fuctional is
\begin{equation}
    J = \text{mass} + \text{gravity} + \text{penalty}.
\end{equation}
The default value of the length scale $\ell$ in the penalty term is defined as:
\begin{equation}
    \ell = 2\max\{\text{cell diameter}, 5h\}
\end{equation}
but users can adjust this to the value of their choice.
This penalty term smooths over numerical artifacts, especially near the ice margins and termini.
In these regions the shallow ice approximation is less applicable, so the error in solving a different set of equations is small compared to the inherent modeling error in using these equations in the first place.

The shallow ice approximation applies well in ice-sheet interiors.
For this reason, and because the equations are particularly simple to solve, this approximation has been a common choice for ice-sheet modeling \citep{kirchner2016shallow}.
Caution must be taken, however, because areas of the ice sheet that experience high vertical velocities (e.g., ice divides) or plug flow (e.g., ice streams) are not well represented by this approximation.

@@

Several studies have compared the shallow stream approximation to 3D models such as Blatter-Pattyn or full Stokes \citep{pattyn2013grounding}.
The most appreciable difference between lower- and higher-order models occurs near the glacier grounding line, where the full Stokes equations can represent bridging stresses \citep{van2013fundamentals}.
The lack of vertical strain rates in the ice viscosity in 2D models can also lead to different equilibrium grounding line positions under the same external forcing.
The shallow stream approximation is warranted for studying individual fast-flowing glaciers where the concern is more with studying processes and responses than with complete accuracy.

@@

The first-order model in icepack is described in the class \texttt{HybridModel}.
The shallow ice and shallow stream models follow directly from the variational principles described above.
The hybrid flow model, while also based on a variational principle, uses two more tricks: terrain-following coordinates and vertical spectral methods.
Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband and not 3D models.
\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem, but this model used a finite difference discretization in the vertical direction.

\paragraph{Terrain-following coordinates}

Rather than use the usual Cartesian coordinate system, the hybrid flow model uses terrain-following coordinates.
The terrain-following vertical coordinate $\zeta$ is
\begin{equation}
    \zeta = \frac{z - b}{h}
\end{equation}
where $b$ is the ice base.
We can then think of the computational domain as the Cartesian product of a 2D footprint domain $\Omega$ and the unit interval $[0, 1]$.

@@

\paragraph{Spectral discretization}

Terrain-following coordinates open up several choices for how to describe the vertical variation of the velocity field.
In Elmer/Ice, for example, the user can extend the finite element discretization into a number of vertical layers.
The number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.

@@

The combination of using extruded meshes and tensor product elements in the vertical direction is referred to as \emph{semi-discretization}.
This approach can be thought of merely as a way to discretize a PDE that has special structure.
Alternatively, we can view vertical semi-discretization as defining a family of models indexed by the number of vertical basis functions.
The order-$d$ model defines a coupled system of PDEs for $d$ vector fields.
Each vector field represents one mode of vertical variability, similar to the distinction between barotropic and baroclinic modes in atmospheric physics and oceanography.
The system is then discretized in the horizontal and solved numerically.
In any case, the code is the same regardless of how one views the underlying mathematics.

The user then has to decide how many vertical modes are enough.
Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
The degree 2 model is the most minimalistic model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
Going to higher degree gives a more accurate approximation at the expense of greater computational effort.

@@

Many aspects of glacier physics are not completely understood.
For example, the most common description of basal friction assumes that glacier sliding occurs through regelation, in which case the basal shear stress can be written
\begin{equation}
    \tau_b = -C|u|^{\frac{1}{m} - 1}u
    \label{eq:weertman-sliding}
\end{equation}
with $m = 3$ \citep{weertman1957sliding}.
Many authors have argued that, in fast-flowing regions of the ice sheet, glacier sliding occurs instead by plastic deformation within the subglacial sediments \citep{tulaczyk2000basal}.
For plastic sliding, the shear stress is dependent on the yield strength of the subglacial sediments and not on the sliding speed, in which case $m = \infty$ would be most appropriate.
Finally, the Schoof sliding law aims to reconcile the two modes of sliding \citep{schoof2005effect}.
Below a critical sliding speed, the shear stress is aymptotic to $|u|^{1/m}$ as in the Weertman sliding law, but above the critical speed the shear stress is independent of the sliding speed.

@@

One of our goals with icepack is to facilitate experimentation with the model physics, even for novice users.
Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would meet these needs.
To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function, which is undesirable.
In C++, one could avoid changing the signature of the diagnostic solve routine by (1) using variadic templates, (2) wrapping the inputs in a class, or (3) passing all arguments in a dictionary.
Using variadic templates or wrapping the inputs in a class would require users to know more about generic or object-oriented programming than a novice might.
Using a dictionary data structure to pass arguments is relatively easier but would be more idiomatic in Python than in C++.

@@

In step 3, any fields that were unnecessary for calculating a given component are ignored.
For example, the gravitational driving power routine will pull out the velocity, thickness, and surface elevation.
It will have access to other fields, for example the ice fluidity.
These extra arguments are ignored.

@@

One key limitation of our approach is that, while users can change what goes into one or more terms of the action functional, they cannot add or remove terms.
This case would call for writing a new diagnostic model.

@@

We have not implemented a model for the surface or englacial transport of meltwater, although it would be possible to include this process.
Instead, we rely on an external scheme to act as a sink of enthalpy for meltwater fraction values above some user-defined critical value, which defaults to 1\%.

@@

Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
The class \texttt{InverseProblem} represents the specification of an inverse problem, which requires:
\begin{itemize}
    \item the model object and the method that solves the diagnostic equation,
    \item the objective and regularization functionals,
    \item the observed field and the name of the argument to the diagnostic solver,
    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver,
    \item extra data passed to the diagnostic solver such as boundary conditions.
\end{itemize}
This class currently assumes that the observed state is always the ice velocity.
In principle the same design would suffice for more complicated inverse problems and this is an area of active development.

The inverse problem class is flexible enough to account for users defining their own parameterization for the rheology or friction coefficient as described in \S\ref{sec:physics-substitution}.
This flexibility with respect to parameterization is not just convenient but essential for common data assimilation workflows.
Nearly all studies in the literature introduce a parameterization of the field to be estimated in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
For example, one could define the friction coefficient $C$ in terms of an auxiliary variable $\beta$ as $C = \beta^2$ to guarantee positivity.
One could also just as easily use $C \propto \exp(\beta)$.
The data assimilation routines in icepack can work the same for any parameterization of the physics because Firedrake provides a rich set of routines for symbolically calculating functional derivatives.
The inverse problem class only needs to know which functional needs to be differentiated with respect to which field.

@@

\appendix
\section{}    %% Appendix A

\subsection{}     %% Appendix A1, A2, etc.

>>>>

\Author[1]{Daniel}{Shapero}
\Author[2]{Jessica}{Badgeley}
\Author[2]{Andrew}{Hoffmann}
\Author[1]{Ian}{Joughin}

@@

\runningauthor{Shapero, Badgeley, Hoffmann, Joughin}

@@

\begin{abstract}
We introduce a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
Icepack is built on the finite element modeling library Firedrake, which uses the domain-specific language UFL for describing weak forms of partial differential equations.
The diagnostic models in icepack are formulated through action principles that are specified in UFL.
The components of each action functional can be substituted for different forms of the user's choosing, which makes it easy to experiment with the model physics.
The action functional itself can be used to define a solver convergence criterion that is independent of the mesh and requires little tuning on the part of the user.
Icepack includes the 2D shallow ice and shallow stream models.
We have also defined a 3D hybrid model based on spectral semi-discretization of the Blatter-Pattyn equations.
Finally, icepack includes a Gauss-Newton solver for inverse problems that runs substantially faster than the BFGS method often used in the glaciological literature.
The overall design philosophy of icepack is to be as usable as possible for a wide a swath of the glaciological community, including both experts and novices in computational science.
\end{abstract}

@@

Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several software packages for glacier flow modeling already exist and are effective in the hands of experts.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item projecting future glacier extent and estimating the sea-level rise contribution from glacier dynamics,
    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood,
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like.
\end{enumerate}
To accomplish these tasks, these tools are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
Many glaciologists receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
In this paper, we introduce a new Python software package for glacier flow modeling called \emph{iceapck}.
Our goal is to make a tool that will be both accessible to novices and more productive for experts.

The glacier flow modeling package closest in spirit to icepack is VarGlaS \citep{brinkerhoff2013data}.
VarGlaS is implemented using the finite element modeling (FEM) package FEniCS \citep{logg2012automated}, whereas icepack is built on top of Firedrake, which began as an outgrowth of FEniCS.
Both packages share a similar goal of saving users from manually writing low-level code for assembling the systems of equations that discretize their model physics.
Instead, users describe the weak form of the partial differential equations they wish to solve using a high-level domain-specific language (DSL) called the \emph{Unified Form Language} or UFL \citep{alnaes2014unified}.
This DSL is embedded entirely into Python, i.e. the complete syntax of UFL can be mapped directly onto overloaded operators in the Python programming language.
Both FEniCS and Firedrake then generate optimized C or C++ code to assemble the discretized system of equations from this symbolic description of the problem \citep{kirby2006compiler, rathgeber2016firedrake}.
Combining a DSL and a code generator frees users from the very error-prone process of writing these assembly kernels themselves, and makes the syntax of the code align more closely to the syntax of written mathematical expressions.
Additionally, having a high-level symbolic description of the problem makes it possible to automatically derive tangent linear models and adjoints \citep{mitusch2019dolfin}.

Icepack improves upon the groundwork laid in VarGlaS in three main respects.
First, icepack includes a simple 3D flow model that uses several features only available in Firedrake: extruded meshes and tensor product finite elements \citep{bercea2016structure, mcrae2016automated} (see \S\ref{sec:physics-hybrid-model}).
Second, icepack's architecture is designed to make it easy for users to alter the various physics components, such as the rheology and basal friction, of any of its constituent models (see \S\ref{sec:physics-substitution}).
Finally, the inverse solver in icepack uses the Gauss-Newton method, which converges faster and more reliably than the BFGS method used in VarGlaS (see section \S\ref{sec:numerics-inverse-solvers}).

@@

The two main components of a glacier flow model are a \emph{diagnostic} and a \emph{prognostic} equation.
The diagnostic equation prescribes the ice velocity through a time-independent, nonlinear, elliptic partial differential equation (PDE).
The prognostic equation prescribes how the ice thickness evolves through conservation of mass.
Mathematically, these two coupled PDEs can be thought of as a differential-algebraic equation.

The rheology and friction coefficient are functions of other fields that have their own evolution equations.
For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock \citep{cuffey2010physics}.
The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.

@@

The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
The prognostic model is a conservative advection equation:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.

The surface elevation is calculated as
\begin{equation}
    s = \max\{b + h, (1 - \rho_I / \rho_W)h\},
\end{equation}
the first case corresponding to grounded ice and the second case corresponding to floating ice.
In the immediate vicinity of the grounding line, the assumption that the floating ice is in hydrostatic equilibrium with the ocean fails.
Most models assume hydrostasy and icepack does as well.
Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.
We have not implemented this feature but support for non-hydrostatic ice shelves may come in a future release of icepack.

@@

There are four \emph{diagnostic} or \emph{momentum transport} models implemented in icepack.
For each of the diagnostic models, we use a formulation of the physics based on a \emph{minimization} or \emph{action principle} \citep{dukowicz2010consistent}.
Action principles are completely equivalent to the usual weak form of a partial differential equation but have certain numerical advantages, as described below.

The most complex and most physically accurate model that we implement is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
(In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
The only approximation that the first-order model makes is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.

From the first-order model, two approximations are possible.
First, the \emph{shallow ice approximation} (SIA) comes from assuming that vertical shear is the dominant mode of ice flow.
The SIA model \citep{hutter1981effect} describes the interior of ice sheets well, where flow can be simply described as bed-parallel shear and surface and basal slopes are small.
This approximation breaks down near ice margins, where basal sliding can be a large fraction of the surface speed and where membrane stresses are substantial.
Second, one could assume that ice flow is purely by horizontal extension and that the surface and bed velocities are practically the same, which is called the \emph{shallow stream approximation} (SSA).
The SSA model describes the fast-flowing margins of the ice sheet best, encompassing grounded ice streams or floating ice shelves \citep{macayeal1989large}.

All the diagnostic models inherit a fundamental nonlinearity in the mechanics of ice flow.
For a Newtonian viscous fluid, the stress tensor $\tau$ and the strain rate tensor $\dot\varepsilon$ are linearly proportional to each other.
Glaciers, however, have a nonlinear constitutive relation:
\begin{equation}
    \dot\varepsilon = A|\tau|^{n - 1}\tau
\end{equation}
where $A$ is the \emph{fluidity} and $n$ is the \emph{Glen flow law exponent} \citep{cuffey2010physics}.
The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.
Strictly speaking, the fluidity should be a tensor field, but is almost universally treated as a scalar.

@@

The action principle is especially useful for designing a robust numerical solver.
For viscous flow problems near to steady-state, the action is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge from any initial guess.
Minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation while having no additional computational cost.
Both formulations are mathematically equivalent.

@@

The shallow ice model can be derived from the Blatter-Pattyn approximation by assuming ice flow is dominated by bed-parallel shear and that surface and basal slopes are small, i.e. $\partial u/\partial x \ll \partial u/\partial z$.
The result is a 2D system of equations for depth-averaged horizontal velocity.
The class in icepack that represents this physics model is called \texttt{ShallowIce}.
The terms in the action functional are:
\begin{align}
    \text{mass} & =\frac{1}{2}\int_\Omega u\cdot u\,dx\\\
    \text{gravity} & = \int_\Omega\frac{2A(\rho_I g)^n}{n+2} h^{n + 1}|\nabla s|^{n - 1}\nabla s\cdot u\,dx\\
    \text{penalty} & = \frac{1}{2}\int_\Omega \ell^2\nabla u\cdot \nabla u\,dx
\end{align}
and the action functional is
\begin{equation}
    J = \text{mass} + \text{gravity} + \text{penalty}.
\end{equation}
The default value of the length scale $\ell$ in the penalty term is defined as:
\begin{equation}
    \ell = 2\max\{\text{cell diameter}, 5h\}
\end{equation}
but users can adjust this to the value of their choice.
This penalty term smooths over numerical artifacts, especially near the ice margins and termini.
In these regions the shallow ice approximation is less applicable, so the error in solving a different set of equations is small compared to the inherent modeling error in using these equations in the first place.

The shallow ice approximation applies well in ice-sheet interiors.
For this reason, and because the equations are particularly simple to solve, this approximation has been a common choice for ice-sheet modeling \citep{cuffey2010physics, kirchner2016shallow}.
This approximation does not work well in areas of the ice sheet where the flow has substantial membrane stresses, like fast outlet glaciers.

@@

Several studies have compared the shallow stream approximation to 3D models such as Blatter-Pattyn or full Stokes \citep{pattyn2013grounding}.
The most appreciable difference between lower- and higher-order models occurs near the glacier grounding line, where the full Stokes equations can represent bridging stresses \citep{van2013fundamentals}.
The lack of vertical strain rates in the ice viscosity in 2D models can also lead to different equilibrium grounding line positions under the same external forcing.
The advantage of the SSA is that that it can capture most of the overall flow features of an individual fast-flowing glacier but at much lower computational cost than the Stokes equations.

@@

The first-order model in icepack is described in the class \texttt{HybridModel}.
The shallow ice and shallow stream models follow directly from the variational principles described above.
The hybrid flow model, while also based on a variational principle, uses two more advanced mathematical techniques: terrain-following coordinates and spectral discretization in the vertical dimension.
Each of these methods has appeared in the literature on glacier modeling before but never all in the same place for a 3D model.
\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband models.
\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem, but this model used a finite difference discretization in the vertical direction.

\textbf{Terrain-following coordinates}.
Rather than the usual Cartesian coordinate system, the hybrid flow model uses terrain-following coordinates.
The terrain-following vertical coordinate $\zeta$ is
\begin{equation}
    \zeta = \frac{z - b}{h}
\end{equation}
where $b$ is the ice base.
We can then think of the computational domain as the Cartesian product of a 2D footprint domain $\Omega$ and the unit interval $[0, 1]$.

@@

\textbf{Spectral discretization}.
Terrain-following coordinates open up several choices for how to describe the vertical variation of the velocity field.
In Elmer/Ice, for example, the user can extend the finite element discretization into a number of vertical layers.
The number of vertical layers is a user-tuneable parameter, depending on the desired resolution along this axis \citep{gagliardini2013capabilities}.

@@

The combination of using extruded meshes and tensor product elements in the vertical direction can be thought of merely as a way to discretize a PDE that has special structure.
Alternatively, we can view discretization in the vertical as defining a family of models indexed by the number of vertical basis functions.
The order-$d$ model defines a coupled system of PDEs for $d$ vector fields.
Each vector field represents one mode of vertical variability, similar to the distinction between barotropic and baroclinic modes in atmospheric physics and oceanography.
The system is then discretized in the horizontal and solved numerically.
In any case, the code is the same regardless of how one views the underlying mathematics.

The user then has to decide how many vertical modes are sufficient.
Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
The degree 2 model is the most minimalistic model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
Going to higher degree gives a more accurate approximation at the expense of greater computational effort.

@@

Many aspects of glacier physics are not completely understood.
For example, the most commonly used sliding law is the power law
\begin{equation}
    \tau_b = -C|u|^{\frac{1}{m} - 1}u
    \label{eq:weertman-sliding}
\end{equation}
for some exponent $m$.
Older research assumed $m = 3$ based on the theory of regelation \citep{weertman1957sliding}, which has since been referred to as the Weertman sliding law.
When $m = \infty$, the basal shear stress is independing of the sliding speed; this is referred to as perfect plasticity.
More recently, \citet{schoof2005effect} proposed the functional form
\begin{equation}
    \tau_b = -C\left(\frac{|u|}{|u| + u_c}\right)^{\frac{1}{m}}\frac{u}{|u|}
\end{equation}
where $u_c$ is some critical speed.
This sliding law acts like the Weertman law when $|u| \ll u_c$ and like the plastic law when $|u| \gg u_c$.
This latter equation has been found to agree best with laboratory experiments \citep{zoet2020slip}.

@@

One of our goals with icepack is to facilitate experimentation with the model physics, even for novice users.
Of the programming languages that are commonly used for scientific computing, only Python and possibly Julia would appear meet these needs.
To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress, which Python accomodates easily.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function, which is undesirable.
In C++, one could avoid changing the signature of the diagnostic solve routine by (1) using variadic templates, (2) wrapping the inputs in a class, or (3) passing all arguments in a dictionary.
Using variadic templates or wrapping the inputs in a class would require users to know more about generic or object-oriented programming than a novice might.
Using a dictionary data structure to pass arguments is relatively easier but would be more idiomatic in Python than in C++.

@@

In step 3, any fields that were unnecessary for calculating a given component are ignored.
For example, the gravitational driving power routine will pull out the velocity, thickness, and surface elevation.
While the routine will not use them, it also has access to other fields, for example the ice fluidity.

@@

@@

We have not implemented a model for the surface or englacial transport of meltwater, although it would be possible to include this process.
Instead, we rely on an external scheme to act as a sink of enthalpy for meltwater fraction values above some user-defined critical value.
A common choice for this critical value is 1\% \citep{aschwanden2012enthalpy}.

@@

Icepack includes a set of routines for estimating the basal friction or rheology coefficients from observational data.
The class \texttt{InverseProblem} represents the specification of an inverse problem, which requires:
\begin{itemize}
    \item the model object and the method that solves the diagnostic equation,
    \item the objective and regularization functionals,
    \item the observed field and the name of the argument to the diagnostic solver,
    \item an initial guess for the field to be estimated and the name of the argument to the diagnostic solver, and
    \item extra data passed to the diagnostic solver such as boundary conditions.
\end{itemize}
This class currently assumes that the observed state is always the ice velocity.
In principle the same design would suffice for more complicated inverse problems and this is an area of active development.

The inverse problem class is flexible enough to account for users defining their own parameterization for the rheology or friction coefficient as described in \S\ref{sec:physics-substitution}.
This flexibility with respect to parameterization is not just convenient but essential for common data assimilation workflows.
Nearly all studies in the literature introduce a parameterization of the field to be estimated in terms of some auxiliary field in order to guarantee positivity \citep{macayeal1992basal, joughin2009basal}.
For example, one could define the friction coefficient $C$ in terms of an auxiliary variable $\beta$ as $C = \beta^2$ to guarantee positivity.
One could also just as easily use $C \propto \exp(\beta)$.
The data assimilation routines in icepack can work the same way for any parameterization of the physics because Firedrake provides a rich set of routines for symbolically calculating functional derivatives.
The inverse problem class only needs to know which functional needs to be differentiated with respect to which field.

@@

\appendix
% \section{}    %% Appendix A

% \subsection{}     %% Appendix A1, A2, etc.
